An Analysis of Performance Bottlenecks in MRI Pre-Processing

### 2024-03-27 paper/paper.tex

\title{An Analysis of Performance Bottlenecks in MRI Pre-Processing}
\author{Mathieu Dugr\'e, Yohan Chatelain, Tristan Glatard}

@@

\begin{abstract}
	MRI pre-processing is a critical step for neuroimaging analysis. However, the computational cost of MRI pre-processing pipelines is a major bottleneck for large cohort studies and some clinical applications. While HPCs and, more recently, Deep Learning have been adopted to accelerate the computations, these techniques require costly hardware and are not accessible to all researchers. Therefore, it is critical to understand the performance bottleneck of MRI pre-processing pipelines to improve their performance. Using Intel VTune profiler, we characterize the bottlenecks of several commonly used MRI-preprocessing pipelines from the ANTs, FSL, and FreeSurfer toolboxes. First, we confirmed that few functions contribute to most of the CPU time. Second, we found that linear interpolation is the main bottleneck for the applications. Data access is also  a substantial bottleneck for the pipelines. Last, we found a bug in the ITK library that impact the performance of ANTs pipeline in single-precision and a potential issue with the OpenMP scaling in FreeSurfer recon-all. Our results provide a reference for future efforts to optimize MRI pre-processing workflows.
\end{abstract}

\section{Introduction}
% Computational bottleneck
Pre-processing of Magnetic Resonance Imaging (MRI) data is a critical step in neuroimaging. Although established pre-processing pipelines exist, they commonly require expensive computation and produce large output and intermediate data. This has been a major bottleneck to study large subject cohorts in neuroimaging. Moreover, the execution time of pipelines can hinder clinical applications where timely data analysis is needed. For those reasons, the neuroimaging community is constantly seeking ways to improve the performance of MRI pre-processing pipelines.

% HPCs
Researcher broadly adopted High Performance Computing (HPC) to accelerate the MRI processing of large cohorts. HPC systems provide a large number of CPU cores and high-bandwidth memory to process subjects in parallel. However, HPCs might not be available to researchers in developing country. Similarly, cloud computing allows scaling the number of core and memory to process large cohort. However, cloud computing requires expertise to setup and can be costly for large cohort. Moreover, data transfer can be slow and might be impossible due to ethic concerns related to human data. In both cases, the benefits from parallelization are limited by the parallel efficiency of the pipeline and the number of subjects to process. Therefore, it can lead to under-utilization of the available resources.

% DL
In recent years, Deep Learning (DL) has shown promising results in MRI pre-preprocessing. For example, tools such as SynthMorph~\cite{Hoffmann2022-hu}, SynthSeg~\cite{Billot2023-vp}, SynthSR~\cite{Iglesias2023-bp}, and SynthStrip~\cite{Hoopes2022-ms} were integrated into FreeSurfer~\cite{Fischl2012-cx} with the effort to provide Deep Learning alternative to traditional algorithms. The main benefit of those tools is the faster runtime they provide. Moreover, A study~\cite{Pepe2023-dm} showed that the results generated by SynthMorph are numerical more stable than those from FreeSurfer. However, training a DL model requires large amount of data and specialized hardware, and is computationally expensive. Additionally, the results from DL models can be challenging to interpret, explaining the slow adoption of DL in MRI pre-processing.

% Alternative solutions
While those techniques can provide significant performance improvements, their additional hardware requirement can be costly and unaccessible to some researchers. Alternatively, improved algorithm designs, compression techniques, or reduced- and mixed-precision techniques could provide substantial performance improvements. However, their implementation can require substantial effort. Therefore, it is critical to understand the performance bottleneck of MRI pre-processing pipelines to improve their performance.

% - No comprehensive study of the performance bottleneck of MRI pre-processing.
The neuroimaging community is well aware that MRI pre-precessing pipelines require long computation time. However, to the best of our knowledge, there is no comprehensive study of the performance bottlenecks in MRI pre-processing. In this paper, we characterize computational cost of several commonly adopted MRI pre-processing pipelines. The results of our analysis serve as a reference for future efforts to optimize MRI pre-processing workflows.

% - Focus on MRI pre-processing.
Magnetic Resonance Imaging (MRI) is a standard tool used by neuroscientists to perform clinical diagnosis and for researchers to develop a better understanding of the brain. Three main MRI techniques exist: structural MRI (sMRI), functional MRI (fMRI), and diffusion MRI (dMRI). While other modalities such as electroencephalogram (EEG), computed tomography (CT), and positron emission tomography (PET) exist, we focus on MRIs for their broad adoption and non-invasive property. Moreover, MRI data analysis is challenging due to computationally expensive pre-processing and large output and intermediate data size.

@@

% - data: OpenNeuro ds004513
We use the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}. We chose this dataset for its wide range of age and equal distribution of biological sex. Using a diverse dataset should help to prevent findings of potential performance bottleneck associated to a specific population.

% - VTune profiler
To profile the pipelines, we use the Intel VTune profiler~\cite{vtune-profiler}. The VTune profiler is a multi-language profiler with low performance overhead and provides information of functions at runtime level. We profile the application using two threading configuration: single-threaded and 32 threads. We aggregate subjects' results for each pipeline to obtain an average performance profile. We compiled each application with debug information to obtain human readable information on function and module names in VTune.

\section{Materials \& Methods}
\subsection{Pipelines}
In this paper, we study the performance bottleneck of fMRIPrep; a commonly used fMRI and sMRI pre-processing workflow. First, we profile the entire workflow to identify the coarse grain performance bottleneck to obtain a coarse view of the workflow's performance. Following, to characterize the performance at a finer grain, we profile the sub-workflows of fMRIPrep: ANTs brainExtraction, ANTs registrationSyN~\cite{Avants2008-ea}, FreeSurfer recon-all~\cite{Dale1999-wu}, FSL FAST~\cite{Zhang2001-hx}, FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}, and FSL MCFLIRT~\cite{Jenkinson2002-od}. We omitted some sub-workflows of fMRIPrep as they were not compatible with our dataset. Moreover, the statistical analysis of the fMRIPrep workflow are not computationally expensive and are not the focus of this paper.

>>>>

\title{An Analysis of Performance Bottlenecks in MRI Pre-Processing}
\author{Mathieu Dugr\'e, Yohan Chatelain, Tristan Glatard\\Department of Computer Science and Software Engineering, Concordia University, Montreal, QC, Canada}

@@

\begin{abstract}
	Magnetic Resonance Image (MRI) pre-processing is a critical step for neuroimaging analysis. However, the computational cost of MRI pre-processing pipelines is a major bottleneck for large cohort studies and some clinical applications. While High-Performance Computing (HPC) and, more recently, Deep Learning have been adopted to accelerate the computations, these techniques require costly hardware and are not accessible to all researchers. Therefore, it is critical to understand the performance bottlenecks of MRI pre-processing pipelines to improve their performance. Using Intel VTune profiler, we characterize the bottlenecks of several commonly used MRI-preprocessing pipelines from the ANTs, FSL, and FreeSurfer toolboxes. We found that few functions contribute to most of the CPU time, and that linear interpolation is the largest contributor. Data access is also  a substantial bottleneck. Last, we identified a bug in the ITK library that impact the performance of ANTs pipeline in single-precision and a potential issue with the OpenMP scaling in FreeSurfer recon-all. Our results provide a reference for future efforts to optimize MRI pre-processing workflows.
\end{abstract}

\section{Introduction}
% Computational bottleneck
Pre-processing of Magnetic Resonance Imaging (MRI) data is a critical step in neuroimaging. Although established pre-processing pipelines exist, they commonly require expensive computation and produce large output and intermediate data. This has been a major bottleneck to study large subject cohorts in neuroimaging. Moreover, the execution time of pipelines can hinder clinical applications where timely data analysis is needed. Therefore, the neuroimaging community is constantly seeking ways to improve the performance of MRI pre-processing pipelines.

% HPCs
\TG{Paragraphs 2 to 5 could be condensed in a more concise way.}
Researchers broadly adopted High Performance Computing (HPC) to accelerate the MRI processing of large cohorts. HPC systems provide a large number of CPU cores and high-bandwidth memory to process subjects in parallel. However, HPCs might not be available to researchers in developing countries \TG{revise that}. Similarly, cloud computing allows scaling the number of core and memory to process large cohort. However, cloud computing requires expertise to setup and can be costly for large cohorts. Moreover, data transfer can be slow and might be impossible due to ethic concerns related to human data. In both cases, the benefits from parallelization are limited by the parallel efficiency of the pipeline and the number of subjects to process. Therefore, it can lead to under-utilization of the available resources.

% DL
In recent years, Deep Learning (DL) has shown promising results in MRI pre-preprocessing. For example, tools such as SynthMorph~\cite{Hoffmann2022-hu}, SynthSeg~\cite{Billot2023-vp}, SynthSR~\cite{Iglesias2023-bp}, and SynthStrip~\cite{Hoopes2022-ms} were integrated into FreeSurfer~\cite{Fischl2012-cx} to provide Deep Learning alternatives to traditional algorithms. The main benefit of those tools is the faster runtime they provide.
% Moreover, a study~\cite{Pepe2023-dm} showed that the results generated by SynthMorph are numerical more stable than those from FreeSurfer.
However, training a DL model requires large amount of data and specialized hardware, and is computationally expensive. Additionally, the results from DL models can be challenging to interpret, explaining the slow adoption of DL in MRI pre-processing.

% Alternative solutions
While those techniques can provide significant performance improvements, their additional hardware requirements can be costly and unaccessible to some researchers. Alternatively, improved algorithm designs, compression techniques, or reduced- and mixed-precision techniques could provide substantial performance improvements. However, their implementation can require substantial effort. Therefore, it is critical to understand the performance bottleneck of MRI pre-processing pipelines to improve their performance.

% - No comprehensive study of the performance bottleneck of MRI pre-processing.
The neuroimaging community is well aware that MRI pre-processing pipelines require long computation time. However, to the best of our knowledge, there is no comprehensive study of the performance bottlenecks in MRI pre-processing. In this paper, we characterize the computational cost of several commonly adopted MRI pre-processing pipelines. The results of our analysis serve as a reference for future efforts to optimize MRI pre-processing workflows.

% - Focus on MRI pre-processing.
MRI is a standard tool used by neuroscientists to perform clinical diagnosis and for researchers to develop a better understanding of the brain. Three main MRI modalities exist: structural MRI (sMRI), functional MRI (fMRI), and diffusion MRI (dMRI). While other modalities such as electroencephalography (EEG), computed tomography (CT), and positron emission tomography (PET) exist, we focus on MRIs for their broad adoption and non-invasiveness. Moreover, MRI data analysis is challenging due to computationally expensive pre-processing and large output and intermediate data size.

@@

% - data: OpenNeuro ds004513
We use the OpenNeuro ds004513 v1.0.2 dataset~\cite{ds004513:1.0.2}. We chose this dataset for its wide range of age and equal distribution of biological sex. Using a diverse dataset should help to prevent findings of potential performance bottleneck associated to a specific population. \TG{Move this to methods}

% - VTune profiler
To profile the pipelines, we use the Intel VTune profiler~\cite{vtune-profiler}. The VTune profiler is a multi-language profiler with low performance overhead and provides information of functions at runtime level. We profile the application using two threading configuration: single-threaded and 32 threads \TG{add justification}. We aggregate subjects' results for each pipeline to obtain an average performance profile. We compiled each application with debug information to obtain human readable information on function and module names in VTune.

\section{Materials \& Methods}
\subsection{Pipelines}
We study the performance bottleneck of fMRIPrep; a commonly used fMRI and sMRI pre-processing workflow. First, we profile the entire workflow to identify the coarse grain performance bottleneck to obtain a coarse view of the workflow's performance. Following, to characterize the performance at a finer grain, we profile the sub-workflows of fMRIPrep: ANTs brainExtraction, ANTs registrationSyN~\cite{Avants2008-ea}, FreeSurfer recon-all~\cite{Dale1999-wu}, FSL FAST~\cite{Zhang2001-hx}, FSL FLIRT~\cite{Jenkinson2002-od,Jenkinson2001-eu,Greve2009-dw}, and FSL MCFLIRT~\cite{Jenkinson2002-od}. We omitted some sub-workflows of fMRIPrep as they were not compatible with our dataset. Moreover, the statistical analysis of the fMRIPrep workflow are not computationally expensive and are not the focus of this paper.
